---
title: "PA1_template"
author: "Andy Vallebueno"
date: "15 de noviembre de 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Activity Monitoring Data Assignment

### 1. Loading and preprocessing the data
We will start off by loading the data and transforming it into a format more suitable for analysis. 

```{r data}
data<-read.csv("activity.csv")
data$date<-as.Date(data$date)
```

### 2. Mean total number of steps taken per day

We begin by calculating the total number of steps taken per day in the following manner. 

```{r totalstepsperday}
sumperday<-with(data, tapply(steps, date, sum, na.rm=TRUE))
sumperday<-as.data.frame(sumperday)
head(sumperday)
```

We now make a histogram of the total number of steps per day and include the mean and median of the total number of steps per day. 

```{r histstepsperday}
hist(sumperday$sumperday, xlab = "Total steps per day", main = "Histogram of the number of steps taken every day")
meansteps = mean(sumperday$sumperday)
mediansteps = median(sumperday$sumperday)
abline(v = meansteps, col = "blue")
abline(v = mediansteps, col = "red")
```

As seen in the plot above, the mean number of steps is `r meansteps` (blue) and the median number of steps is `r mediansteps` (red).

### 3. Average daily activity pattern
We now plot the 5-minute interval and the average number of steps taken, averaged across all days. 

```{r averageactivity}
intervals<-unique(data$interval)
intervalaverage<-with(data, tapply(steps, interval, mean, na.rm = TRUE))
intervalaverage<-as.data.frame(intervalaverage)
intervalaverage$intervals<-intervals
plot(intervalaverage$intervals, intervalaverage$intervalaverage, type = "l", main = "Average steps taken per interval across all days", xlab = "Interval", ylab = "Number of steps")
maxint<-which.max(intervalaverage$intervalaverage)
maximumint<-intervalaverage[maxint,2]
```

The 5-minute interval which on average across all the days in the dataset contains the maximum number of steps is `r maximumint`.

### 4. Imputing missing values
We start off by calculating the total number of missing values in the dataset. 

```{r missingvalues}
missingvalues<-is.na(data$steps)
totalmissingvalues<-sum(missingvalues)
```

In total, we have `r totalmissingvalues` missing values in the dataset. 

To fill these missing values, we will use the average steps across all days for the specific interval of the missing value, and create a new dataset equal to the original dataset but with the missing data filled in. 

```{r newdataset}
datanew<-data
vectorlength<-1:length(datanew$steps)

for (i in vectorlength) {
  if (missingvalues[i] == TRUE) {
    nainterval<-data$interval[i]
    subsetinterval<-subset(intervalaverage, intervals == nainterval, intervalaverage)
    datanew$steps[i]<-subsetinterval[1,1]
  } else {
    datanew$steps[i]<- data$steps[i]
  }
}

head(datanew)
```

We now make a new histogram of the total number of steps taken each day and calculare the mean and median. 

```{r newhiststepsperday}
sumperdaynew<-with(datanew, tapply(steps, date, sum))
sumperdaynew<-as.data.frame(sumperdaynew)
hist(sumperdaynew$sumperdaynew, xlab = "Total steps per day", main = "Histogram of the number of steps taken every day")
meanstepsnew = mean(sumperdaynew$sumperdaynew)
medianstepsnew = median(sumperdaynew$sumperdaynew)
abline(v = meanstepsnew, col = "blue")
abline(v = medianstepsnew, col = "red")
```

The new mean number of steps per day is `r meanstepsnew` while the new median steps per day is `r medianstepsnew`. The difference in the mean is `r meanstepsnew - meansteps` and the difference in the median is `r medianstepsnew - mediansteps`. The impact of imputing missing data on the estimates of the total daily number of steps is an `r ((meanstepsnew/meansteps) - 1)*100 `% increase in the mean and an `r ((medianstepsnew/mediansteps) - 1)*100 `% increase in the median. 

### 5. Activity patterns between weekdays and weekends
To determine whether there are activity patterns between weekdays and weekends, we first create a new factor variable in the new dataset with two levels - "weekday" and "weekend". 

```{r weekdaysweekends}
datanew$day<-0

for (i in vectorlength) {
  day<-weekdays(datanew$date[i])
  weekdaysdays<-c("lunes", "martes", "miércoles", "jueves", "viernes")
  if (day %in% weekdaysdays == TRUE) {
    datanew$day[i]<-"Weekday"
  } else {
    datanew$day[i]<-"Weekend"
  }
}
datanew$day<-as.factor(datanew$day)
head(datanew)
```

We now make a panel plot containing a time series plot of the 5-minute interval axis and the average number of steps taken, averaged across all weekday days or weekend days. 

```{r plotweekdays}
library(ggplot2)
intervalsweekdays<-with(datanew, aggregate(steps ~ interval + day, datanew, mean))

g<- ggplot(intervalsweekdays, aes(interval, steps))
g<- g + geom_line() + facet_grid(day~.) + ggtitle("Average number of steps taken across all days pery 5-minute interval") + labs(x = "Interval", y = "Number of steps")
print(g)
```

